---
title: "Population Health Impact and Economic Consequences of Storms"
author: "Ben Kesseler"
date: "May 22, 2016"
output: html_document
---

<style type="text/css">

th, td {  /* Table  */
   font-size: 10px;
}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***

# Synopsis
This report examines the U.S. National Oceanic and Atmospheric Administration's 
(NOAA) storm database, comparing economic damages, injuries, and fatalities. 
All analysis was performed at the national level, ignoring specific locations. 
Individual "events" are not easily determined to assess impacts per "event", so 
impacts are considered in aggregate over time.

The most damaging and dangerous event types are tornados, hurricanes/typhoons, 
thunderstorm winds, lightning, hail, floods, storm surges/tides, rip currents, 
and heat.

Heat (combining heat and excessive heat) is the most dangerous event in terms of
fatalities, and second most dangerous in terms of injuries, while causing little
economic damage, so it should be prepared for in warm climates.

Tornados are very damaging in both health and economic terms, and should be
prepared for in areas prone to experience them.

Floods are extremely damaging economically, while also impacting health fairly
severely, so flood prone regions should take relevant precautions.

Hurricanes and typhoons, perhaps because of the advanced warning usually 
available, cause less impact to health, while causing the second highest 
economic damages.

***

# Data Processing

## Loading the Raw Data

Library packages are loaded for the entire analysis. Code output has been
suppressed for this code only.

```{r libraries, results = 'hide', message = FALSE}
libraries <- c("ggplot2",
               "knitr",
               "tools",
               "readr",
               "scales",
               "lubridate",
               "pander",
               "reshape",
               "ggrepel",
               "gridExtra"
               )
sapply(libraries, library, character.only = TRUE)
```

We start with the [NOAA Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)
that is archived on the course website and documented [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) and [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf).

```{r dataimport, cache = TRUE}
source_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
target_local_archive <- "storm_data.csv.bz2"

if (!file.exists(target_local_archive)) {
  download.file(source_url,target_local_archive)
}

storm_data <- read_csv(target_local_archive, na = "NA")
```

***

## Initial Cleaning

Each row is a very detailed description of a location, event type, and damage,
if any. Actual logical events such as a regional storm or other weather impact 
might entail many rows, and there is no obvious way to usefully aggregate the 
rows to logical events.

The analysis covers the entire United States, so location is not important.

As a result, the data will be aggregated by event type at the year level. This 
will allow the analysis to calculate the impact at an annual level if needed,
and across the entire time period for the main analysis.

Once the data is loaded from the .csv.bz2 file, I keep only the year, event
type, fatalities, injuries, property damage, and crop damage.

I convert the economic impacts into a consistent scale. I'm unsure of the 
meaning of non-standard exponents in the data, so I'm treating them as 1, after 
documenting their prevalence.

```{r datacleaning}
storm_data_cleaned <- storm_data[, c("BGN_DATE",
                                     "EVTYPE",
                                     "FATALITIES",
                                     "INJURIES",
                                     "PROPDMG",
                                     "PROPDMGEXP",
                                     "CROPDMG",
                                     "CROPDMGEXP"
                                     )
                                 ]

colnames(storm_data_cleaned) <- c("BGN_DATE",
                                  "event_type",
                                  "fatalities",
                                  "injuries",
                                  "property_damage",
                                  "property_damage_exponent",
                                  "crop_damage",
                                  "crop_damage_exponent"
                                  )

storm_data_cleaned$year <- year(as.POSIXct(storm_data_cleaned$BGN_DATE,
                                           format = "%m/%d/%Y"))

mappable_exponents_property <- sum(storm_data_cleaned$property_damage_exponent == "B") +
                               sum(storm_data_cleaned$property_damage_exponent == "M") +
                               sum(storm_data_cleaned$property_damage_exponent == "K") +
                               sum(storm_data_cleaned$property_damage_exponent == "")

mappable_exponents_crop <- sum(storm_data_cleaned$crop_damage_exponent == "B") +
                           sum(storm_data_cleaned$crop_damage_exponent == "M") +
                           sum(storm_data_cleaned$crop_damage_exponent == "K") +
                           sum(storm_data_cleaned$crop_damage_exponent == "")

percent_unmappable_property <- 1 - mappable_exponents_property / nrow(storm_data_cleaned)

percent_unmappable_crop <- 1 - mappable_exponents_crop / nrow(storm_data_cleaned)
```

So `r percent(percent_unmappable_property)` of property damage records have
unmappable exponents, and `r percent(percent_unmappable_crop)` of crop damage
records have unmappable exponents. That's a very small percent of the records, 
so treating them as 1 should be reasonable.

***

I now convert things to a consistent numeric scale, so summation functions will
work properly.

```{r datacleaning2}
scalar_mapping <- data.frame("property_damage_exponent" = c("B","M","K"),
                             "property_damage_numeric_exponent" = c(1000000000,
                                                                    1000000,
                                                                    1000),
                             "crop_damage_exponent" = c("B","M","K"),
                             "crop_damage_numeric_exponent" = c(1000000000,
                                                                1000000,
                                                                1000)
                            )

storm_data_cleaned <- merge(storm_data_cleaned,
                            scalar_mapping[, 
                                           c("property_damage_exponent",
                                             "property_damage_numeric_exponent"
                                            )
                                          ],
                            all.x = TRUE)
storm_data_cleaned$property_damage_numeric_exponent[
  is.na(storm_data_cleaned$property_damage_numeric_exponent)
  ] <- 1

storm_data_cleaned <- merge(storm_data_cleaned,
                            scalar_mapping[, 
                                           c("crop_damage_exponent",
                                             "crop_damage_numeric_exponent"
                                            )
                                          ],
                            all.x = TRUE)
storm_data_cleaned$crop_damage_numeric_exponent[
  is.na(storm_data_cleaned$crop_damage_numeric_exponent)
  ] <- 1

storm_data_cleaned$property_damage <- storm_data_cleaned$property_damage * 
                                      storm_data_cleaned$property_damage_numeric_exponent

storm_data_cleaned$crop_damage <- storm_data_cleaned$crop_damage * 
                                  storm_data_cleaned$crop_damage_numeric_exponent

storm_data_cleaned$economic_damage <- storm_data_cleaned$property_damage + 
                                      storm_data_cleaned$crop_damage
```

## Data Aggregation

Now I will summarize the data at the level needed for plotting and further
exploration - annual by event type.

```{r dataaggregation}
storm_data_cleaned <- storm_data_cleaned[, c("year",
                                             "event_type",
                                             "fatalities",
                                             "injuries",
                                             "economic_damage")]

storm_data_summary <- aggregate(. ~ year + event_type,
                                data = storm_data_cleaned,
                                FUN = sum)

storm_data_count <- aggregate(fatalities ~ year + event_type,
                              data = storm_data_cleaned,
                              FUN = length)

storm_data_summary <- cbind(storm_data_summary,storm_data_count$fatalities)

storm_data_summary$count <- storm_data_count$fatalities

count_by_year <- aggregate(count ~ year, data = storm_data_summary, FUN = sum)

count_by_year_table <- data.frame("year1" = count_by_year[1:16, 1], 
                                  "event_count1" = count_by_year[1:16, 2],
                                  "year2" = count_by_year[17:32, 1],
                                  "event_count2" = count_by_year[17:32, 2],
                                  "year3" = count_by_year[33:48, 1],
                                  "event_count3" = count_by_year[33:48, 2],
                                  "year4" = c(count_by_year[49:62, 1], "", ""),
                                  "event_count4" = c(count_by_year[49:62, 2], "", "")
)

panderOptions("table.alignment.default","left")
panderOptions("table.split.table",Inf)
pander(count_by_year_table)
```

It looks like there is a marked increase in the number of events in 1995. I am
concerned that data from previous years doesn't accurately represent the
distribution of event types, so I am restricting the analysis to 1995 and
forward.

## Event Type Clean-up

Here I will summarize the data by event_type, for the years 1995 and after, and
I will remove event types that have no fatalities, injuries, or economic damage,
and begin cleaning event types.

```{r datasubsetting}
storm_data_summary <- storm_data_summary[storm_data_summary$year >= 1995, ]

storm_data_summary <- aggregate(. ~ event_type,
                                data = storm_data_summary[, c("event_type",
                                                              "count",
                                                              "fatalities",
                                                              "injuries",
                                                              "economic_damage")
                                                          ],
                                FUN = sum)

storm_data_summary <- storm_data_summary[storm_data_summary$fatalities > 0 |
                                         storm_data_summary$injuries > 0 |
                                         storm_data_summary$economic_damage > 0
                                         , ]

storm_data_summary$event_type <- tolower(storm_data_summary$event_type)

storm_data_summary$event_type <- gsub("[^a-zA-Z0-9]","",storm_data_summary$event_type)

storm_data_summary <- aggregate(. ~ event_type,
                                data = storm_data_summary,
                                FUN = sum)

storm_data_summary$event_match[storm_data_summary$event_type == "astronomicallowtide"]    <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "avalanche"]              <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "blizzard"]               <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "coastalflood"]           <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "coldwindchill"]          <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "debrisflow"]             <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "densefog"]               <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "densesmoke"]             <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "drought"]                <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "dustdevil"]              <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "duststorm"]              <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "excessiveheat"]          <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "extremecoldwindchill"]   <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "flashflood"]             <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "flood"]                  <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "frostfreeze"]            <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "funnelcloud"]            <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "freezingfog"]            <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "hail"]                   <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "heat"]                   <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "heavyrain"]              <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "heavysnow"]              <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "highsurf"]               <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "highwind"]               <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "hurricanetyphoon"]       <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "icestorm"]               <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "lakeeffectsnow"]         <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "lakeshoreflood"]         <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "lightning"]              <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "marinehail"]             <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "marinehighwind"]         <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "marinestrongwind"]       <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "marinethunderstormwind"] <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "ripcurrent"]             <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "seiche"]                 <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "sleet"]                  <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "stormsurgetide"]         <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "strongwind"]             <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "thunderstormwind"]       <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "tornado"]                <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "tropicaldepression"]     <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "tropicalstorm"]          <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "tsunami"]                <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "volcanicash"]            <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "waterspout"]             <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "wildfire"]               <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "winterstorm"]            <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "winterweather"]          <- 1

storm_data_summary$event_match[is.na(storm_data_summary$event_match)] <- 0

initial_matching <- aggregate(. ~ event_match,
                              data = storm_data_summary[, -1],
                              FUN = sum)

percent_unmatched_count <- initial_matching$count[initial_matching$event_match == 0] / 
  sum(initial_matching$count)

percent_unmatched_injuries <- initial_matching$injuries[initial_matching$event_match == 0] / 
  sum(initial_matching$injuries)

percent_unmatched_fatalities <- initial_matching$fatalities[initial_matching$event_match == 0] / 
  sum(initial_matching$fatalities)

percent_unmatched_economic_damage <- initial_matching$economic_damage[initial_matching$event_match == 0] / 
  sum(initial_matching$economic_damage)
```

After minimal text formatting, I find that when comparing to the expected 48
event types, I'm unable to match `r percent(percent_unmatched_count)`
of the non-zero events, which account for `r percent(percent_unmatched_economic_damage)`
of the economic damage, `r percent(percent_unmatched_injuries)` of the injuries,
and `r percent(percent_unmatched_fatalities)` of the fatalities.

This clearly needs some work.

***

What event types remain?

``` {r cleanup}
event_type_list <- data.frame("event_type" = storm_data_summary[1:77, 1],
                              "event_type2" = storm_data_summary[78:154, 1],
                              "event_type3" = storm_data_summary[155:231, 1],
                              "event_type4" = c(storm_data_summary[232:307, 1], ""))

panderOptions("table.alignment.default", "left")
panderOptions("table.split.table",Inf)
pander(event_type_list)
```

***

Now I'm going to clean up some of the obvious event_types errors.

```{r textreplacement}
storm_data_summary$event_type[grep("^coastalflood", storm_data_summary$event_type)] <- "coastalflood"
storm_data_summary$event_type[storm_data_summary$event_type == "erosioncstlflood"] <- "coastalflood"
storm_data_summary$event_type[storm_data_summary$event_type == "tidalflooding"] <- "coastalflood"

storm_data_summary$event_type[storm_data_summary$event_type == "extendedcold"] <- "coldwindchill"
storm_data_summary$event_type[grep("^cold", storm_data_summary$event_type)] <- "coldwindchill"
storm_data_summary$event_type[storm_data_summary$event_type == "recordcold"] <- "coldwindchill"

storm_data_summary$event_type[storm_data_summary$event_type == "fog"] <- "densefog"

storm_data_summary$event_type[storm_data_summary$event_type == "droughtexcessiveheat"] <- "drought"

storm_data_summary$event_type[storm_data_summary$event_type == "dustdevilwaterspout"] <- "dustdevil"

storm_data_summary$event_type[storm_data_summary$event_type == "extremecold"] <- "extremecoldwindchill"
storm_data_summary$event_type[storm_data_summary$event_type == "extremewindchill"] <- "extremecoldwindchill"

storm_data_summary$event_type[grep("^flashflood", storm_data_summary$event_type)] <- "flashflood"
storm_data_summary$event_type[grep("^floodflash", storm_data_summary$event_type)] <- "flashflood"
storm_data_summary$event_type[storm_data_summary$event_type == "rapidlyrisingwater"] <- "flashflood"

storm_data_summary$event_type[grep("^flood", storm_data_summary$event_type)] <- "flood"
storm_data_summary$event_type[storm_data_summary$event_type == "breakupflooding"] <- "flood"
storm_data_summary$event_type[storm_data_summary$event_type == "riverandstreamflood"] <- "flood"
storm_data_summary$event_type[grep("^riverflood", storm_data_summary$event_type)] <- "flood"
storm_data_summary$event_type[grep("^urbanflood", storm_data_summary$event_type)] <- "flood"
storm_data_summary$event_type[storm_data_summary$event_type == "urbanandsmallstreamfloodin"] <- "flood"
storm_data_summary$event_type[storm_data_summary$event_type == "urbansmallstreamflood"] <- "flood"
storm_data_summary$event_type[storm_data_summary$event_type == "urbansmlstreamfld"] <- "flood"
storm_data_summary$event_type[storm_data_summary$event_type == "mudslidesurbanflooding"] <- "flood"
storm_data_summary$event_type[storm_data_summary$event_type == "ruralflood"] <- "flood"

storm_data_summary$event_type[grep("freeze$", storm_data_summary$event_type)] <- "frostfreeze"
storm_data_summary$event_type[grep("frost$", storm_data_summary$event_type)] <- "frostfreeze"

storm_data_summary$event_type[grep("^hail", storm_data_summary$event_type)] <- "hail"
storm_data_summary$event_type[storm_data_summary$event_type == "gustywindhail"] <- "hail"
storm_data_summary$event_type[storm_data_summary$event_type == "smallhail"] <- "hail"

storm_data_summary$event_type[grep("^heat", storm_data_summary$event_type)] <- "heat"
storm_data_summary$event_type[storm_data_summary$event_type == "recordheat"] <- "heat"
storm_data_summary$event_type[storm_data_summary$event_type == "warmweather"] <- "heat"

storm_data_summary$event_type[grep("^heavyrain", storm_data_summary$event_type)] <- "heavyrain"
storm_data_summary$event_type[storm_data_summary$event_type == "hvyrain"] <- "heavyrain"
storm_data_summary$event_type[storm_data_summary$event_type == "excessiverainfall"] <- "heavyrain"
storm_data_summary$event_type[storm_data_summary$event_type == "torrentialrainfall"] <- "heavyrain"

storm_data_summary$event_type[grep("^heavysnow", storm_data_summary$event_type)] <- "heavysnow"
storm_data_summary$event_type[grep("^snow", storm_data_summary$event_type)] <- "heavysnow"
storm_data_summary$event_type[storm_data_summary$event_type == "excessivesnow"] <- "heavysnow"

storm_data_summary$event_type[grep("^heavysurf", storm_data_summary$event_type)] <- "highsurf"
storm_data_summary$event_type[grep("^highsurf", storm_data_summary$event_type)] <- "highsurf"
storm_data_summary$event_type[storm_data_summary$event_type == "hazardoussurf"] <- "highsurf"

storm_data_summary$event_type[storm_data_summary$event_type == "heavyswells"] <- "highswells"

storm_data_summary$event_type[grep("^highwind", storm_data_summary$event_type)] <- "highwind"
storm_data_summary$event_type[grep("^gustywind", storm_data_summary$event_type)] <- "highwind"

storm_data_summary$event_type[grep("^hurricane", storm_data_summary$event_type)] <- "hurricanetyphoon"
storm_data_summary$event_type[storm_data_summary$event_type == "typhoon"] <- "hurricanetyphoon"

storm_data_summary$event_type[storm_data_summary$event_type == "icejamfloodminor"] <- "icejamflooding"
storm_data_summary$event_type[storm_data_summary$event_type == "iceonroad"] <- "icyroads"
storm_data_summary$event_type[storm_data_summary$event_type == "iceroads"] <- "icyroads"

storm_data_summary$event_type[storm_data_summary$event_type == "lakeflood"] <- "lakeshoreflood"

storm_data_summary$event_type[storm_data_summary$event_type == "landslides"] <- "landslide"

storm_data_summary$event_type[storm_data_summary$event_type == "lightsnowfall"] <- "lightsnow"

storm_data_summary$event_type[grep("^lightning", storm_data_summary$event_type)] <- "lightning"
storm_data_summary$event_type[storm_data_summary$event_type == "ligntning"] <- "lightning"

storm_data_summary$event_type[storm_data_summary$event_type == "marinetstmwind"] <- "marinethunderstormwind"

storm_data_summary$event_type[storm_data_summary$event_type == "mixedprecip"] <- "mixedprecipitation"

storm_data_summary$event_type[storm_data_summary$event_type == "mudslides"] <- "mudslide"

storm_data_summary$event_type[grep("^ripcurrent", storm_data_summary$event_type)] <- "ripcurrent"

storm_data_summary$event_type[storm_data_summary$event_type == "stormsurge"] <- "stormsurgetide"

storm_data_summary$event_type[storm_data_summary$event_type == "strongwinds"] <- "strongwind"
storm_data_summary$event_type[storm_data_summary$event_type == "stormforcewinds"] <- "strongwind"
storm_data_summary$event_type[storm_data_summary$event_type == "gustywinds"] <- "strongwind"
storm_data_summary$event_type[storm_data_summary$event_type == "icestrongwinds"] <- "strongwind"

storm_data_summary$event_type[grep("^thunderstorm", storm_data_summary$event_type)] <- "thunderstormwind"
storm_data_summary$event_type[grep("^tstmwind", storm_data_summary$event_type)] <- "thunderstormwind"
storm_data_summary$event_type[grep("^severethunderstorm", storm_data_summary$event_type)] <- "thunderstormwind"
storm_data_summary$event_type[storm_data_summary$event_type == "thunerstormwinds"] <- "thunderstormwind"
storm_data_summary$event_type[storm_data_summary$event_type == "thunderstormwins"] <- "thunderstormwind"
storm_data_summary$event_type[storm_data_summary$event_type == "tunderstormwind"] <- "thunderstormwind"
storm_data_summary$event_type[storm_data_summary$event_type == "thundeerstormwinds"] <- "thunderstormwind"
storm_data_summary$event_type[storm_data_summary$event_type == "thunderestormwinds"] <- "thunderstormwind"
storm_data_summary$event_type[storm_data_summary$event_type == "thundertormwinds"] <- "thunderstormwind"

storm_data_summary$event_type[grep("^tornado", storm_data_summary$event_type)] <- "tornado"

storm_data_summary$event_type[grep("^tropicalstorm", storm_data_summary$event_type)] <- "tropicalstorm"

storm_data_summary$event_type[storm_data_summary$event_type == "unseasonablecold"] <- "unseasonablycold"

storm_data_summary$event_type[storm_data_summary$event_type == "waterspouttornado"] <- "waterspout"

storm_data_summary$event_type[grep("^wildforestfire", storm_data_summary$event_type)] <- "wildfire"
storm_data_summary$event_type[grep("fires$", storm_data_summary$event_type)] <- "wildfire"
storm_data_summary$event_type[storm_data_summary$event_type == "brushfire"] <- "wildfire"

storm_data_summary$event_type[grep("^wind", storm_data_summary$event_type)] <- "wind"

storm_data_summary$event_type[storm_data_summary$event_type == "winterstormhighwinds"] <- "winterstorm"

storm_data_summary$event_type[grep("^winterweather", storm_data_summary$event_type)] <- "winterweather"
storm_data_summary$event_type[storm_data_summary$event_type == "wintrymix"] <- "winterweather"

storm_data_summary <- aggregate(. ~ event_type,
                                data = storm_data_summary,
                                FUN = sum)
```

Things are a bit more clear now. Let's see how these align to the 48 event types
now.

***

```{r sortto48}
storm_data_summary$event_match[storm_data_summary$event_type == "astronomicallowtide"]    <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "avalanche"]              <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "blizzard"]               <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "coastalflood"]           <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "coldwindchill"]          <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "debrisflow"]             <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "densefog"]               <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "densesmoke"]             <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "drought"]                <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "dustdevil"]              <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "duststorm"]              <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "excessiveheat"]          <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "extremecoldwindchill"]   <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "flashflood"]             <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "flood"]                  <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "frostfreeze"]            <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "funnelcloud"]            <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "freezingfog"]            <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "hail"]                   <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "heat"]                   <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "heavyrain"]              <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "heavysnow"]              <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "highsurf"]               <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "highwind"]               <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "hurricanetyphoon"]       <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "icestorm"]               <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "lakeeffectsnow"]         <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "lakeshoreflood"]         <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "lightning"]              <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "marinehail"]             <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "marinehighwind"]         <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "marinestrongwind"]       <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "marinethunderstormwind"] <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "ripcurrent"]             <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "seiche"]                 <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "sleet"]                  <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "stormsurgetide"]         <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "strongwind"]             <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "thunderstormwind"]       <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "tornado"]                <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "tropicaldepression"]     <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "tropicalstorm"]          <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "tsunami"]                <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "volcanicash"]            <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "waterspout"]             <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "wildfire"]               <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "winterstorm"]            <- 1
storm_data_summary$event_match[storm_data_summary$event_type == "winterweather"]          <- 1

final_matching <- aggregate(. ~ event_match,
                            data = storm_data_summary[, -1],
                            FUN = sum)

percent_unmatched_count_final <- final_matching$count[final_matching$event_match == 0] / 
  sum(final_matching$count)

percent_unmatched_injuries_final <- final_matching$injuries[final_matching$event_match == 0] / 
  sum(final_matching$injuries)

percent_unmatched_fatalities_final <- final_matching$fatalities[final_matching$event_match == 0] / 
  sum(final_matching$fatalities)

percent_unmatched_economic_damage_final <- final_matching$economic_damage[final_matching$event_match == 0] / 
  sum(final_matching$economic_damage)
```

After more aggressive remapping, I find that when comparing to the expected 48
event types, I'm unable to match `r percent(percent_unmatched_count_final)`
of the non-zero events, which account for `r percent(percent_unmatched_economic_damage_final)`
of the economic damage, `r percent(percent_unmatched_injuries_final)` of the injuries,
and `r percent(percent_unmatched_fatalities_final)` of the fatalities.

That's an acceptable error rate.

A little text adjustment for the final plotting and analysis section.

``` {r finaltextnaming}
storm_data_summary$event_type[storm_data_summary$event_type == "astronomicallowtide"] <- "Astronomical Low Tide"
storm_data_summary$event_type[storm_data_summary$event_type == "avalanche"] <- "Avalanche"
storm_data_summary$event_type[storm_data_summary$event_type == "blizzard"] <- "Blizzard"
storm_data_summary$event_type[storm_data_summary$event_type == "coastalflood"] <- "Coastal Flood"
storm_data_summary$event_type[storm_data_summary$event_type == "coldwindchill"] <- "Cold/Wind Chill"
storm_data_summary$event_type[storm_data_summary$event_type == "debrisflow"] <- "Debris Flow"
storm_data_summary$event_type[storm_data_summary$event_type == "densefog"] <- "Dense Fog"
storm_data_summary$event_type[storm_data_summary$event_type == "densesmoke"] <- "Dense Smoke"
storm_data_summary$event_type[storm_data_summary$event_type == "drought"] <- "Drought"
storm_data_summary$event_type[storm_data_summary$event_type == "dustdevil"] <- "Dust Devil"
storm_data_summary$event_type[storm_data_summary$event_type == "duststorm"] <- "Dust Storm"
storm_data_summary$event_type[storm_data_summary$event_type == "excessiveheat"] <- "Excessive Heat"
storm_data_summary$event_type[storm_data_summary$event_type == "extremecoldwindchill"] <- "Extreme Cold/Wind Chill"
storm_data_summary$event_type[storm_data_summary$event_type == "flashflood"] <- "Flash Flood"
storm_data_summary$event_type[storm_data_summary$event_type == "flood"] <- "Flood"
storm_data_summary$event_type[storm_data_summary$event_type == "frostfreeze"] <- "Frost/Freeze"
storm_data_summary$event_type[storm_data_summary$event_type == "funnelcloud"] <- "Funnel Cloud"
storm_data_summary$event_type[storm_data_summary$event_type == "freezingfog"] <- "Freezing Fog"
storm_data_summary$event_type[storm_data_summary$event_type == "hail"] <- "Hail"
storm_data_summary$event_type[storm_data_summary$event_type == "heat"] <- "Heat"
storm_data_summary$event_type[storm_data_summary$event_type == "heavyrain"] <- "Heavy Rain"
storm_data_summary$event_type[storm_data_summary$event_type == "heavysnow"] <- "Heavy Snow"
storm_data_summary$event_type[storm_data_summary$event_type == "highsurf"] <- "High Surf"
storm_data_summary$event_type[storm_data_summary$event_type == "highwind"] <- "High Wind"
storm_data_summary$event_type[storm_data_summary$event_type == "hurricanetyphoon"] <- "Hurricane (Typhoon)"
storm_data_summary$event_type[storm_data_summary$event_type == "icestorm"] <- "Ice Storm"
storm_data_summary$event_type[storm_data_summary$event_type == "lakeeffectsnow"] <- "Lake-Effect Snow"
storm_data_summary$event_type[storm_data_summary$event_type == "lakeshoreflood"] <- "Lakeshore Flood"
storm_data_summary$event_type[storm_data_summary$event_type == "lightning"] <- "Lightning"
storm_data_summary$event_type[storm_data_summary$event_type == "marinehail"] <- "Marine Hail"
storm_data_summary$event_type[storm_data_summary$event_type == "marinehighwind"] <- "Marine High Wind"
storm_data_summary$event_type[storm_data_summary$event_type == "marinestrongwind"] <- "Marine Strong Wind"
storm_data_summary$event_type[storm_data_summary$event_type == "marinethunderstormwind"] <- "Marine Thunderstorm Wind"
storm_data_summary$event_type[storm_data_summary$event_type == "ripcurrent"] <- "Rip Current"
storm_data_summary$event_type[storm_data_summary$event_type == "seiche"] <- "Seiche"
storm_data_summary$event_type[storm_data_summary$event_type == "sleet"] <- "Sleet"
storm_data_summary$event_type[storm_data_summary$event_type == "stormsurgetide"] <- "Storm Surge/Tide"
storm_data_summary$event_type[storm_data_summary$event_type == "strongwind"] <- "Strong Wind"
storm_data_summary$event_type[storm_data_summary$event_type == "thunderstormwind"] <- "Thunderstorm Wind"
storm_data_summary$event_type[storm_data_summary$event_type == "tornado"] <- "Tornado"
storm_data_summary$event_type[storm_data_summary$event_type == "tropicaldepression"] <- "Tropical Depression"
storm_data_summary$event_type[storm_data_summary$event_type == "tropicalstorm"] <- "Tropical Storm"
storm_data_summary$event_type[storm_data_summary$event_type == "tsunami"] <- "Tsunami"
storm_data_summary$event_type[storm_data_summary$event_type == "volcanicash"] <- "Volcanic Ash"
storm_data_summary$event_type[storm_data_summary$event_type == "waterspout"] <- "Waterspout"
storm_data_summary$event_type[storm_data_summary$event_type == "wildfire"] <- "Wildfire"
storm_data_summary$event_type[storm_data_summary$event_type == "winterstorm"] <- "Winter Storm"
storm_data_summary$event_type[storm_data_summary$event_type == "winterweather"] <- "Winter Weather"
```

***

# Results

## Fatality, Injury, and Economic Damage Sources

I need to examine the distribution of injuries, fatalaties, and economic damage
by the event types present in this data.

I'll be using top 20 for most of these analyses, to make the plots easier to 
review.

```{r figure1prep}
storm_data_final <- storm_data_summary[storm_data_summary$event_match == 1, ]
storm_data_final$event_type <- factor(storm_data_final$event_type)

storm_data_final$combined_health <- storm_data_final$injuries + 
                                    storm_data_final$fatalities

storm_data_final$event_type <- factor(storm_data_final$event_type,
                                      levels = storm_data_final$event_type[
                                        order(storm_data_final$combined_health,
                                              decreasing = TRUE)])

storm_data_final_plot1 <- head(storm_data_final[
  order(-storm_data_final$combined_health), ], 20)

storm_data_final_plot1 <- melt(data = storm_data_final_plot1[, c("event_type",
                                                                 "fatalities",
                                                                 "injuries")
                                                            ], id="event_type")

storm_data_final_plot1$variable <- as.factor(toTitleCase(
    as.character(storm_data_final_plot1$variable)))

First_Plot <- ggplot(data = storm_data_final_plot1,
                       aes(x = event_type, y = value, fill = variable)) +
                geom_bar(stat = "identity", position = "dodge") +
                theme_bw() +
                coord_flip() +
                ggtitle("Injuries & Fatalities by Event Type") +
                xlab("Event Types") +
                ylab("Injuries & Fatalities") + 
                scale_fill_manual(values = c(Fatalities = "red", Injuries = "orange")) +
                theme(legend.position = c(0.9, 0.8), 
                      legend.title = element_blank(),
                      panel.grid.major.y = element_line(colour = "#808080"))

storm_data_final_plot2 <- head(storm_data_final[
  order(-storm_data_final$combined_health), ], 20)

Second_Plot <- ggplot(data = storm_data_final_plot2, aes(x = injuries, 
                                                         y = fatalities)) + 
  geom_point() +
  theme_bw() +
  ggtitle("Injuries vs Fatalities by Event Type") +
  xlab("Injuries") +
  ylab("Fatalities") +
  geom_label_repel(data = subset(storm_data_final_plot2, 
                                 (injuries > 1500 | fatalities > 250)),
                   aes(label = event_type))

storm_data_final_plot3 <- storm_data_final_plot2
storm_data_final_plot3$ratio <- storm_data_final_plot2$fatalities / storm_data_final_plot2$injuries

Third_Plot <- ggplot(data = storm_data_final_plot3, aes(x = event_type, y = ratio, fill = ratio)) + 
  theme_bw() + 
  ggtitle("Fatality to Injury Ratio by Event Type") +
  xlab("Event Types") +
  ylab("Ratio of Fatalities to Injuries") + 
  geom_bar(stat = "identity") + 
  coord_flip() +
  scale_fill_gradient("ratio", low = "orange", high = "red") +
  theme(legend.position = c(0.85, 0.15),
        legend.title = element_blank(),
        legend.direction = "horizontal")
```

## Figure 1 - Injuries & Fatality Exploration

```{r figure1, fig.align = "center", fig.width = 10, fig.height = 10, fig.cap = "The first plot shows injuries and fatalities for the top 20 event types. The second plot shows injuries vs fatalities for the same subset of events. The third plot shows the ratio of fatalities to injuries, to demonstrate which events are the most dangerous, independent of overall prevalence."}
grid.arrange(First_Plot, Second_Plot, Third_Plot, ncol = 1)
```

***

Fatalities and injuries seem to be correlated, although Tornados cause 
signifigantly more injuries than any other event. However, Rip Currents are more
dangerous to life, independent of overall risk of encountering the events.

***

```{r figure2prep}
storm_data_final$event_type <- factor(storm_data_final$event_type,
                                      levels = storm_data_final$event_type[
                                        order(storm_data_final$economic_damage,
                                              decreasing = TRUE)])

storm_data_final_plot4 <- head(storm_data_final[
  order(-storm_data_final$economic_damage), ], 20)

storm_data_final_plot4$economic_damage <- storm_data_final_plot4$economic_damage / 1000000000

Fourth_Plot <- ggplot(data = storm_data_final_plot4,
                       aes(x = event_type, y = economic_damage, fill = economic_damage)) +
                geom_bar(stat = "identity") +
                theme_bw() +
                coord_flip() +
                ggtitle("Economic Damage by Event Type") +
                xlab("Event Types") +
                ylab("Economic Damage in Billions of USD (unadjusted)") + 
                scale_fill_gradient("economic_damage", low = "green", high="dark green") +
                theme(legend.position = c(0.85, 0.9), 
                      legend.direction = "horizontal",
                      legend.title = element_blank(),
                      panel.grid.major.y = element_line(colour = "#808080"))
```

## Figure 2 - Economic Damage Exploration

```{r figure2, fig.align = "center", fig.width = 10, fig.height = 4, fig.cap = "Economic Damage in Billions of USD is displayed, for the top 20 event types. Flood may be somewhat overstated as all flood events must have a damage statement, even if an estimate."}

Fourth_Plot
```

Floods, hurricanes/typhoons, storm surges/tides, and tornados are the most
damaging events, economically.

***

```{r figure3prep}
Fifth_Plot <- ggplot(data = storm_data_final, aes(x = economic_damage/1000000000, 
                                                  y = injuries)) + 
  geom_point() +
  theme_bw() +
  ggtitle("Economic Damage vs Injuries by Event Type") +
  xlab("Economic Damage") +
  ylab("Injuries") +
  geom_label_repel(data = subset(storm_data_final, 
                                 (injuries > 2500 | economic_damage > 15000000000)),
                   aes(label = event_type))

Sixth_Plot <- ggplot(data = storm_data_final, aes(x = economic_damage/1000000000, 
                                                  y = fatalities)) + 
  geom_point() +
  theme_bw() +
  ggtitle("Economic Damage vs Fatalities by Event Type") +
  xlab("Economic Damage") +
  ylab("Fatalities") +
  geom_label_repel(data = subset(storm_data_final, 
                                 (fatalities > 400 | economic_damage > 15000000000)),
                   aes(label = event_type))
```

## Figure 3 - Economic Damage vs Injuries & Fatalities by Event Type

```{r figure3, fig.align = "center", fig.width = 10, fig.height = 4, fig.cap = "The first plot shows economic damages vs injuries for all event types. The second plot compares economics damages to fatalities."}
grid.arrange(Fifth_Plot, Sixth_Plot, ncol = 2)
```

# Further Exploration

This analysis did not cover events over time or by location, while data on both
was available in the raw data. A further analysis could plot injuries, 
fatalities, and economic damage as a density plot over the United States, by
utilizing the latitude and longitude present in the data. Similarly, after 
subsetting to post-1994 data to ensure consistent reporting methods, events
and damages could be compared over time, to determine if any events are occuring
more frequently, or are causing more impact, over time.



